---
title: "Profiling Neuron Inflammation Driven by Alzheimer’s Disease and Type 2 Diabetes Associated Metabolites"
author: "Brendan K. Ball"
date: "21AUG23"
output: html_document
---

## 0. Download the Required Packages
The following packages were used for the RStudio software version 1.4.1717 "Juliet Rose" for macOS.

```{r Library Packages Installation, warning=FALSE}
# Downloads and calls from the library the necessary packages needed to run the code
if(!require(stringr)) install.packages("stringr", repos = "http://cran.us.r-project.org")
suppressWarnings(library (stringr, verbose = FALSE))
if(!require(gtools)) install.packages("gtools", repos = "http://cran.us.r-project.org")
suppressWarnings(library (gtools, verbose = FALSE))
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
suppressWarnings(library (dplyr, verbose = FALSE))
if(!require(factoextra)) install.packages("factoextra", repos = "http://cran.us.r-project.org")
suppressWarnings(library (factoextra, verbose = FALSE))
if(!require(devtools)) install.packages("devtools", repos = "http://cran.us.r-project.org")
suppressWarnings(library (devtools, verbose = FALSE))
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
suppressWarnings(library (data.table, verbose = FALSE))
if(!require(effects)) install.packages("effects", repos = "http://cran.us.r-project.org")
suppressWarnings(library (effects, verbose = FALSE))
if(!require(pheatmap)) install.packages("pheatmap", repos = "http://cran.us.r-project.org")
suppressWarnings(library (pheatmap, verbose = FALSE))
if(!require(mixOmics)) install.packages("mixOmics", repos = "http://cran.us.r-project.org")
suppressWarnings(library (mixOmics, verbose = FALSE))
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
suppressWarnings(library (tidyverse, verbose = FALSE))
if(!require(purrr)) install.packages("purrr", repos = "http://cran.us.r-project.org")
suppressWarnings(library (purrr, verbose = FALSE))
```

## 1. Data Preparation for Analysis
The Observed Concentration (OC) of the cytokines measured from the Luminex assay is used for the data analysis. The imported CSV file is first simplified for easier interpretation of the data and use of code.

```{r Import of Observed Concentration (OC) Results and Prepare Data, warning=FALSE}
# Download the OC cytokine data front the 384-well plate, T2D-AD research work over the summer
OC_cytokine <- read.csv("~/Desktop/20221025_ObservedConc.csv")
# Remove the selected rows
OC_cytokine <- OC_cytokine[-c(1:6,9,72:76),]
# Remove the first two columns
OC_cytokine <- OC_cytokine[-c(1:2)]
# Rename the first row, first column cell as "Description"
OC_cytokine[1,1] <- "Description"
# Pull the column information to use for the column name
OC_colname <- OC_cytokine[1,]
# Rename the columns 
colnames(OC_cytokine) <- OC_colname
# Remove the two rows at the top
OC_cytokine <- OC_cytokine[-c(1:2),]
# Remove the columns that have significant number of missing information
OC_cytokine <- OC_cytokine[-c(3, 4, 6, 13:16, 24)]
# Reset the row number names
rownames(OC_cytokine) <- NULL
# Remove character strings to simplify the data/metabolite name
OC_cytokine$Description <- gsub("Reagent Name: ", "", as.character(OC_cytokine$Description))
OC_cytokine$Description <- gsub("Sample ID: ", "", as.character(OC_cytokine$Description))

# Remove the numbers after the metabolites to cluster easier by name
OC_cytokine[9:62,1] <- gsub("1|2|3", "", as.character(OC_cytokine[9:62,1]))

# Removes all special characters, in this case, the * sign that indicates value extrapolated from standard range
for (i in 2:ncol(OC_cytokine)) {
  OC_cytokine[,i] <- gsub("^\\*", "", as.character(OC_cytokine[,i]))
}

# For loop to make all columns within the OC_cytokine data frame numeric, except for the first column
for (i in 2:ncol(OC_cytokine)) {
  OC_cytokine[,i] <- as.numeric(OC_cytokine[,i])
}
```

### 1.1. Data Processing with Standard Values
The OC cytokine data is processed by replacing any <OOR, or below the lowest level of quantification with the lowest recorded standard value. This is done for any m

```{r Substitute Missing Values with Lowest Standard Measurements}
## Eotaxin
# Replace OOR< value for niacinamide vehicle with the standard 7 value
OC_cytokine[38,2] <- OC_cytokine[7,2]
# Replace OOR< values for D-sorbitol vehicle with the standard 7 value
OC_cytokine[60,2] <- OC_cytokine[7,2]

## IL-1alpha
# Replace OOR< values for arachidonic acid with the standard 8 value 
OC_cytokine[15,5] <- OC_cytokine[8,5]
OC_cytokine[16,5] <- OC_cytokine[8,5]
# Replace OOR< values for retinol vehicle with the standard 8 value
OC_cytokine[31,5] <- OC_cytokine[8,5]
# Replace OOR< values for niacinamide vehicle with the standard 8 value
OC_cytokine[38,5] <- OC_cytokine[8,5]
# Replace OOR< values for L-tryptophan with the standard 8 value
OC_cytokine[39,5] <- OC_cytokine[8,5]
# Replace OOR< values for aminoadipic acid vehicle with the standard 8 value
OC_cytokine[55,5] <- OC_cytokine[8,5]
# Replace OOR< values for D-sorbitol vehicle with the standard 8 value
OC_cytokine[62,5] <- OC_cytokine[8,5]

## IL6
# Replace OOR< values for retinol vehicle with the standard 8 value
OC_cytokine[31,10] <- OC_cytokine[8,10]
# Replace OOR< values for niacinamide vehicle with the standard 8 value
OC_cytokine[36,10] <- OC_cytokine[8,10]
# Replace OOR< values for L-tryptophan with the standard 8 value
OC_cytokine[39,10] <- OC_cytokine[8,10]
# Replace OOR< values for aminoadipic acid vehicle with the standard 8 value
OC_cytokine[54,10] <- OC_cytokine[8,10]
OC_cytokine[56,10] <- OC_cytokine[8,10]
# Replace OOR< values for D-sorbitol with the standard 8 value
OC_cytokine[59,10] <- OC_cytokine[8,10]
# Replace OOR< values for D-sorbitol vehicle with the standard 8 value
OC_cytokine[62,10] <- OC_cytokine[8,10]

## RANTES
# Replace OOR< values for retinol vehicle with the standard 8 value
OC_cytokine[31,16] <- OC_cytokine[8, 16]
```

### 1.2. Log2 Data Normalization Respect to Metabolite Vehicles
A new data frame without the standards is created, and the metabolite names are renamed from the abbreviation to the full name. The processed data is then normalized respect to the reported cytokine concentrations from the vehicle groups. 

```{r Log2 Normalization with the Mean Values}
# Create a new data frame without the standards
OC_dat <- OC_cytokine[c(9:62),]
# Reset the row numbers name
rownames(OC_dat) <- NULL

# Renaming the metabolites to the full name
OC_dat[c(1:3),1] <- "Lauric Acid"
OC_dat[c(4:6),1] <- "Lauric Acid (V)"
OC_dat[c(7:9),1] <- "Arachidonic Acid"
OC_dat[c(10:12),1] <- "Arachidonic Acid (V)"
OC_dat[c(13:15),1] <- "Fructose"
OC_dat[c(16:18),1] <- "Fructose (V)"
OC_dat[c(19:21),1] <- "Retinol"
OC_dat[c(22:24),1] <- "Retinol (V)"
OC_dat[c(25:27),1] <- "Niacinamide"
OC_dat[c(28:30),1] <- "Niacinamide (V)"
OC_dat[c(31:33),1] <- "L-Tryptophan"
OC_dat[c(34:36),1] <- "L-Tryptophan (V)"
OC_dat[c(37:39),1] <- "Asparagine"
OC_dat[c(40:42),1] <- "Asparagine (V)"
OC_dat[c(43:45),1] <- "Aminoadipic Acid"
OC_dat[c(46:48),1] <- "Aminoadipic Acid (V)"
OC_dat[c(49:51),1] <- "D-Sorbitol"
OC_dat[c(52:54),1] <- "D-Sorbitol (V)"

### Taking the mean for each separate cytokines
# Take the vehicles separately in a new data frame for data analysis
OC_dat_veh <- OC_dat[grepl("(V)", OC_dat$Description),] #Vehicles only
# Combine the similar description names (group_by) then take the average for each metabolite OC vehicle group
## Eotaxin
OC_veh_avg_eotaxin <- OC_dat_veh  %>% group_by(Description) %>% summarize(Eotaxin = mean(`Eotaxin (74)`))
OC_veh_avg_eotaxin <- as.data.frame(OC_veh_avg_eotaxin)
## IFN-gamma
OC_veh_avg_IFNgamma <- OC_dat_veh  %>% group_by(Description) %>% summarize(`IFN-gamma` = mean(`IFN-gamma (34)`))
OC_veh_avg_IFNgamma <- as.data.frame(OC_veh_avg_IFNgamma)
## IL-1beta
OC_veh_avg_IL1b <- OC_dat_veh  %>% group_by(Description) %>% summarize(`IL-1beta` = mean(`IL-1Beta (19)`))
OC_veh_avg_IL1b <- as.data.frame(OC_veh_avg_IL1b)
## IL-1alpha
OC_veh_avg_IL1a <- OC_dat_veh  %>% group_by(Description) %>% summarize(`IL-1alpha` = mean(`IL-1alpha (53)`))
OC_veh_avg_IL1a <- as.data.frame(OC_veh_avg_IL1a)
## IL10
OC_veh_avg_IL10 <- OC_dat_veh  %>% group_by(Description) %>% summarize(IL10 = mean(`IL10 (56)`))
OC_veh_avg_IL10 <- as.data.frame(OC_veh_avg_IL10)
## IL12p40
OC_veh_avg_IL12p40 <- OC_dat_veh  %>% group_by(Description) %>% summarize(IL12p40 = mean(`IL12p40 (76)`))
OC_veh_avg_IL12p40 <- as.data.frame(OC_veh_avg_IL12p40)
## IL12p70
OC_veh_avg_IL12p70 <- OC_dat_veh  %>% group_by(Description) %>% summarize(IL12p70 = mean(`IL12p70 (78)`))
OC_veh_avg_IL12p70 <- as.data.frame(OC_veh_avg_IL12p70)
## IL13
OC_veh_avg_IL13 <- OC_dat_veh  %>% group_by(Description) %>% summarize(IL13 = mean(`IL13 (37)`))
OC_veh_avg_IL13 <- as.data.frame(OC_veh_avg_IL13)
## IL6
OC_veh_avg_IL6 <- OC_dat_veh  %>% group_by(Description) %>% summarize(IL6 = mean(`IL6 (38)`))
OC_veh_avg_IL6 <- as.data.frame(OC_veh_avg_IL6)
## IL9
OC_veh_avg_IL9 <- OC_dat_veh  %>% group_by(Description) %>% summarize(IL9 = mean(`IL9 (33)`))
OC_veh_avg_IL9 <- as.data.frame(OC_veh_avg_IL9)
## KC
OC_veh_avg_KC <- OC_dat_veh  %>% group_by(Description) %>% summarize(KC = mean(`KC (57)`))
OC_veh_avg_KC <- as.data.frame(OC_veh_avg_KC)
## MCP-1
OC_veh_avg_MCP1 <- OC_dat_veh  %>% group_by(Description) %>% summarize(`MCP-1` = mean(`MCP-1 (51)`))
OC_veh_avg_MCP1 <- as.data.frame(OC_veh_avg_MCP1)
## MIP-1beta
OC_veh_avg_MIP1b <- OC_dat_veh  %>% group_by(Description) %>% summarize(`MIP-1beta` = mean(`MIP-1Beta (75)`))
OC_veh_avg_MIP1b <- as.data.frame(OC_veh_avg_MIP1b)
## MIP-1alpha
OC_veh_avg_MIP1a <- OC_dat_veh  %>% group_by(Description) %>% summarize(`MIP-1alpha` = mean(`MIP-1alpha (77)`))
OC_veh_avg_MIP1a <- as.data.frame(OC_veh_avg_MIP1a)
## RANTES
OC_veh_avg_RANTES <- OC_dat_veh  %>% group_by(Description) %>% summarize(`RANTES` = mean(`RANTES (55)`))
OC_veh_avg_RANTES <- as.data.frame(OC_veh_avg_RANTES)

# Merge the data sets into a list
OC_avg_list <- list(OC_veh_avg_eotaxin, OC_veh_avg_IFNgamma, OC_veh_avg_IL1b, OC_veh_avg_IL1a, OC_veh_avg_IL10, OC_veh_avg_IL12p40, 
                    OC_veh_avg_IL12p70, OC_veh_avg_IL13, OC_veh_avg_IL6, OC_veh_avg_IL9, OC_veh_avg_KC, OC_veh_avg_MCP1,
                    OC_veh_avg_MIP1b, OC_veh_avg_MIP1a, OC_veh_avg_RANTES)
# Combine the data frames together using the list
OC_avg_all <- Reduce(function(x, y) merge(x, y, all=TRUE), OC_avg_list)

OC_dat_met <- OC_dat[!grepl("(V)", OC_dat$Description),] #Metabolites only
# Reset the row numbers name
rownames(OC_dat_met) <- NULL
# Rename each description to have the replicate number included
OC_dat_met[1,1] <- "Lauric Acid (1)"
OC_dat_met[2,1] <- "Lauric Acid (2)"
OC_dat_met[3,1] <- "Lauric Acid (3)"
OC_dat_met[4,1] <- "Arachidonic Acid (1)"
OC_dat_met[5,1] <- "Arachidonic Acid (2)"
OC_dat_met[6,1] <- "Arachidonic Acid (3)"
OC_dat_met[7,1] <- "Fructose (1)"
OC_dat_met[8,1] <- "Fructose (2)"
OC_dat_met[9,1] <- "Fructose (3)"
OC_dat_met[10,1] <- "Retinol (1)"
OC_dat_met[11,1] <- "Retinol (2)"
OC_dat_met[12,1] <- "Retinol (3)"
OC_dat_met[13,1] <- "Niacinamide (1)"
OC_dat_met[14,1] <- "Niacinamide (2)"
OC_dat_met[15,1] <- "Niacinamide (3)"
OC_dat_met[16,1] <- "L-Tryptophan (1)"
OC_dat_met[17,1] <- "L-Tryptophan (2)"
OC_dat_met[18,1] <- "L-Tryptophan (3)"
OC_dat_met[19,1] <- "Asparagine (1)"
OC_dat_met[20,1] <- "Asparagine (2)"
OC_dat_met[21,1] <- "Asparagine (3)"
OC_dat_met[22,1] <- "Aminoadipic Acid (1)"
OC_dat_met[23,1] <- "Aminoadipic Acid (2)"
OC_dat_met[24,1] <- "Aminoadipic Acid (3)"
OC_dat_met[25,1] <- "D-Sorbitol (1)"
OC_dat_met[26,1] <- "D-Sorbitol (2)"
OC_dat_met[27,1] <- "D-Sorbitol (3)"

# Create a data frame with the computations for averaged vehicles
OC_dat_avg <- OC_dat_met
# Match alphabetical order similar to the structure of the vehicles data 
OC_dat_avg <- OC_dat_avg %>% arrange(Description)
## Take the log2(Metabolite/Vehicle) Calculation for each replicate
# For Aminoadipic Acid replicates 1-3
OC_dat_avg[1,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[1,2:ncol(OC_dat_avg)]/OC_avg_all[1,2:ncol(OC_dat_avg)])
OC_dat_avg[2,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[2,2:ncol(OC_dat_avg)]/OC_avg_all[1,2:ncol(OC_dat_avg)])
OC_dat_avg[3,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[3,2:ncol(OC_dat_avg)]/OC_avg_all[1,2:ncol(OC_dat_avg)])
# For Arachidonic Acid replicates 1-3
OC_dat_avg[4,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[4,2:ncol(OC_dat_avg)]/OC_avg_all[2,2:ncol(OC_dat_avg)])
OC_dat_avg[5,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[5,2:ncol(OC_dat_avg)]/OC_avg_all[2,2:ncol(OC_dat_avg)])
OC_dat_avg[6,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[6,2:ncol(OC_dat_avg)]/OC_avg_all[2,2:ncol(OC_dat_avg)])
# For Asparagine replicates 1-3
OC_dat_avg[7,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[7,2:ncol(OC_dat_avg)]/OC_avg_all[3,2:ncol(OC_dat_avg)])
OC_dat_avg[8,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[8,2:ncol(OC_dat_avg)]/OC_avg_all[3,2:ncol(OC_dat_avg)])
OC_dat_avg[9,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[9,2:ncol(OC_dat_avg)]/OC_avg_all[3,2:ncol(OC_dat_avg)])
# For D-Sorbitol replicates 1-3
OC_dat_avg[10,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[10,2:ncol(OC_dat_avg)]/OC_avg_all[4,2:ncol(OC_dat_avg)])
OC_dat_avg[11,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[11,2:ncol(OC_dat_avg)]/OC_avg_all[4,2:ncol(OC_dat_avg)])
OC_dat_avg[12,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[12,2:ncol(OC_dat_avg)]/OC_avg_all[4,2:ncol(OC_dat_avg)])
# For Fructose replicates 1-3
OC_dat_avg[13,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[13,2:ncol(OC_dat_avg)]/OC_avg_all[5,2:ncol(OC_dat_avg)])
OC_dat_avg[14,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[14,2:ncol(OC_dat_avg)]/OC_avg_all[5,2:ncol(OC_dat_avg)])
OC_dat_avg[15,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[15,2:ncol(OC_dat_avg)]/OC_avg_all[5,2:ncol(OC_dat_avg)])
# For L-Tryptophan replicates 1-3
OC_dat_avg[16,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[16,2:ncol(OC_dat_avg)]/OC_avg_all[6,2:ncol(OC_dat_avg)])
OC_dat_avg[17,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[17,2:ncol(OC_dat_avg)]/OC_avg_all[6,2:ncol(OC_dat_avg)])
OC_dat_avg[18,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[18,2:ncol(OC_dat_avg)]/OC_avg_all[6,2:ncol(OC_dat_avg)])
# For Lauric Acid replicates 1-3
OC_dat_avg[19,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[19,2:ncol(OC_dat_avg)]/OC_avg_all[7,2:ncol(OC_dat_avg)])
OC_dat_avg[20,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[20,2:ncol(OC_dat_avg)]/OC_avg_all[7,2:ncol(OC_dat_avg)])
OC_dat_avg[21,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[21,2:ncol(OC_dat_avg)]/OC_avg_all[7,2:ncol(OC_dat_avg)])
# For Niacinamide replicates 1-3
OC_dat_avg[22,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[22,2:ncol(OC_dat_avg)]/OC_avg_all[8,2:ncol(OC_dat_avg)])
OC_dat_avg[23,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[23,2:ncol(OC_dat_avg)]/OC_avg_all[8,2:ncol(OC_dat_avg)])
OC_dat_avg[24,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[24,2:ncol(OC_dat_avg)]/OC_avg_all[8,2:ncol(OC_dat_avg)])
# For Retinol replicates 1-3
OC_dat_avg[25,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[25,2:ncol(OC_dat_avg)]/OC_avg_all[9,2:ncol(OC_dat_avg)])
OC_dat_avg[26,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[26,2:ncol(OC_dat_avg)]/OC_avg_all[9,2:ncol(OC_dat_avg)])
OC_dat_avg[27,2:ncol(OC_dat_avg)] <- log2(OC_dat_avg[27,2:ncol(OC_dat_avg)]/OC_avg_all[9,2:ncol(OC_dat_avg)])
```

## 2. Interpretation of the Data
The interpretation of the data includes hierarchical clustering with heat maps, principcal component analysis, partial least squared discriminant analysis, and box-and-whisker plots. Statistical analysis includes Kruskal Wallis and Mann Whitney for significance. 

### 2.1. Heat Map for Mean metabolite Values with q-values from Kruskal-Wallis Test
A heat map is generated with the log2 fold change for each cytokine concentration across different metabolite stimulation. The cytokines are hierarchically clustered.

```{r Generating the Heat Map for Mean Metabolite Values, fig.height=4.5, fig.width=5}
# Create a placement variable for the matrix
OC_avg_m <- OC_dat_avg
# Create a variable to hold the row names
OC_avg_rowname <- OC_avg_m$Description
# Remove the "Description" column to move as a row name
OC_avg_matrix <- OC_avg_m[c(2:16)]
# Replace the current row name with the "Description" of the metabolites
rownames(OC_avg_matrix) <- OC_avg_rowname 
# Rename the columns to the cytokines
colnames(OC_avg_matrix)[1] <- "Eotaxin"
colnames(OC_avg_matrix)[2] <- "IFN-gamma"
colnames(OC_avg_matrix)[3] <- "IL-1beta"
colnames(OC_avg_matrix)[4] <- "IL-1alpha"
colnames(OC_avg_matrix)[5] <- "IL10"
colnames(OC_avg_matrix)[6] <- "IL12p40"
colnames(OC_avg_matrix)[7] <- "IL12p70"
colnames(OC_avg_matrix)[8] <- "IL13"
colnames(OC_avg_matrix)[9] <- "IL6"
colnames(OC_avg_matrix)[10] <- "IL9"
colnames(OC_avg_matrix)[11] <- "KC"
colnames(OC_avg_matrix)[12] <- "MCP-1"
colnames(OC_avg_matrix)[13] <- "MIP-1beta"
colnames(OC_avg_matrix)[14] <- "MIP-1alpha"
colnames(OC_avg_matrix)[15] <- "RANTES"

# Creating different groups for the metabolites to reorganize order based on the associations with T2D and AD
OC_matrix_AAA <- OC_avg_matrix[1:3,]
OC_matrix_AA <- OC_avg_matrix[4:6,]
OC_matrix_Asp <- OC_avg_matrix[7:9,]
OC_matrix_DS <- OC_avg_matrix[10:12,]
OC_matrix_F <- OC_avg_matrix[13:15,]
OC_matrix_LT <- OC_avg_matrix[16:18,]
OC_matrix_LA <- OC_avg_matrix[19:21,]
OC_matrix_N <- OC_avg_matrix[22:24,]
OC_matrix_R <- OC_avg_matrix[25:27,]

# Bind the groups of metabolites by respective associations with T2D and AD
# AD protective: LT, N;    T2D: LA, Asp;    AD/T2D protective: R;     AD/T2D: AAA, AA, DS, F
OC_avg_matrix <- rbind(OC_matrix_LT, OC_matrix_N, OC_matrix_LA, OC_matrix_Asp, OC_matrix_R, OC_matrix_AAA, OC_matrix_AA, OC_matrix_DS, OC_matrix_F)

# Make the data frame a matrix
OC_avg_matrix <- as.matrix(OC_avg_matrix)

# Generate the heat map
heatmapfig <- pheatmap(OC_avg_matrix,
  cluster_rows = F,
  cluster_cols = T,
  color = colorRampPalette(c("blue", "white", "red"))(10000),
  border_color = "black",
  breaks = seq(-5, 5, length.out = 10000),
  angle_col = 45,
  fontsize_col = 9,
  fontsize_row = 9,
  cellwidth=20)
```

A Kruskal-Wallis (KW) test is performed across different disease-associations for each metabolite. The p-values for each cytokine across the different disease-association is recorded and used for the heatmap. 

```{r Kruskal-Wallis Test}
# Perform Kruskal-Wallis tests and store p-values in a vector

# Create a KW-defined data frame
OC_condition_KW <- OC_dat_avg

# Replace the description with the condition label
colnames(OC_condition_KW)[1] <- "Condition"

# Renaming the metabolite names to respective condition group
# AD protective: LT, N;    T2D: LA, Asp;    AD/T2D protective: R;     AD/T2D: AAA, AA, DS, F
OC_condition_KW[c(1:3),1] <- "AD & T2D"
OC_condition_KW[c(4:6),1] <- "AD & T2D"
OC_condition_KW[c(7:9),1] <- "T2D"
OC_condition_KW[c(10:12),1] <- "AD & T2D"
OC_condition_KW[c(13:15),1] <- "AD & T2D"
OC_condition_KW[c(16:18),1] <- "AD Protective"
OC_condition_KW[c(19:21),1] <- "T2D"
OC_condition_KW[c(22:24),1] <- "AD Protective"
OC_condition_KW[c(25:27),1] <- "AD & T2D Protective"


p_values <- c()
for (i in 2:ncol(OC_condition_KW)) {
  kwTest2 <- kruskal.test(OC_condition_KW[,i], OC_condition_KW$Condition)
  p_values <- c(p_values, kwTest2$p.value)
}

# Perform Benjamini-Hochberg correction
adjusted_p_values <- p.adjust(p_values, method = "BH")

# Combine the original p-values and adjusted p-values into a data frame
p_value_df <- data.frame(OriginalPValue = p_values, AdjustedPValue = adjusted_p_values)

# Renaming the cytokine column
p_value_df[1,3] <- "Eotaxin"
p_value_df[2,3] <- "IFN-gamma"
p_value_df[3,3] <- "IL-1beta"
p_value_df[4,3] <- "IL-1alpha"
p_value_df[5,3] <- "IL10"
p_value_df[6,3] <- "IL12p40"
p_value_df[7,3] <- "IL12p70"
p_value_df[8,3] <- "IL13"
p_value_df[9,3] <- "IL6"
p_value_df[10,3] <- "IL9"
p_value_df[11,3] <- "KC"
p_value_df[12,3] <- "MCP-1"
p_value_df[13,3] <- "MIP-1beta"
p_value_df[14,3] <- "MIP-1alpha"
p_value_df[15,3] <- "RANTES"

# Rename the columns
colnames(p_value_df)[1] <- "p-value"
colnames(p_value_df)[2] <- "BH p-value"
colnames(p_value_df)[3] <- "Cytokine"

# Reset the row count number
rownames(p_value_df) <- NULL
# Make the p-value column numeric
p_value_df$`p-value` <- as.numeric(p_value_df$`p-value`)
p_value_df$`BH p-value` <- as.numeric(p_value_df$`BH p-value`)
# Prepare a method to color code the graph based on p-value greater or less than 0.05
p_value_df <- p_value_df %>%
  mutate(sign = `BH p-value` <= 0.05)

# Adjust the positions on the bar chart to display bars from lowest p-value to largest value
positions <- c("MIP-1alpha", "IL13", "IFN-gamma", "IL-1alpha", "Eotaxin", "KC", "IL12p70", "RANTES", "MIP-1beta", "IL9", "IL6", "IL10", "IL-1beta", "IL12p40", "MCP-1")

# Plot the KW BH corrected p-value results
ggplot(p_value_df) +
  geom_col(aes(x = `BH p-value`, y = Cytokine, fill=sign), colour = "black", width = 1) +
  ylab(NULL) + xlab("p-value (color cuttoff at p-value > 0.05)") + theme_minimal() + 
  theme(legend.position = "none", text=element_text(color="black", size = 10), 
        axis.text = element_text(color = "black", size=10)) + 
  xlim(c(-0,0.65)) + scale_y_discrete(limits = positions)
```

### 2.2. Principal Component Analysis
A principal component analysis (PCA) is done for the observed concentration groups. Groupings and coloring is based on A) indivudual metabolites and B) disease-associations.

```{r Principal Component Analysis of the Metabolites and Disease-Associated Groups, fig.height=5.2, fig.width=2.8}
# Begin by creating a renamed copy of the data for editing
OC_dat_group <- OC_dat_avg

# Renaming the metabolites to the full name (originally as abbreviated letters)
OC_dat_group[c(1:3),1] <- "Aminoadipic Acid"
OC_dat_group[c(4:6),1] <- "Arachidonic Acid"
OC_dat_group[c(7:9),1] <- "Asparagine"
OC_dat_group[c(10:12),1] <- "D-Sorbitol"
OC_dat_group[c(13:15),1] <- "Fructose"
OC_dat_group[c(16:18),1] <- "L-Tryptophan"
OC_dat_group[c(19:21),1] <- "Lauric Acid"
OC_dat_group[c(22:24),1] <- "Niacinamide"
OC_dat_group[c(25:27),1] <- "Retinol"

# Prepare the principal components
OC_dat_group_pr <- prcomp(OC_dat_group[c(2:ncol(OC_dat_group))], center = FALSE, scale = FALSE) #Metabolites

# Form a PCA by categorizing based on T2D/AD affiliations
PCA_groupings <- as.data.frame(c("AD+T2D", "AD+T2D", "AD+T2D", "AD+T2D", "AD+T2D", "AD+T2D", "T2D", "T2D", "T2D", "AD+T2D", "AD+T2D", "AD+T2D", "AD+T2D", "AD+T2D", "AD+T2D", "AD*", "AD*", "AD*", "T2D", "T2D", "T2D", "AD*", "AD*", "AD*", "AD*+T2D*", "AD*+T2D*", "AD*+T2D*"))
# Rename the column as a conditional group
colnames(PCA_groupings)[1] <- "Condition"

# Combine the groupings column with the original data set
OC_dat_group_cond <- cbind(OC_dat_group, PCA_groupings)

# Prepare the principal components
OC_dat_group_cond_pr <- prcomp(OC_dat_group_cond[c(2:ncol(OC_dat_group))], center = FALSE, scale = FALSE) #Metabolites

# Create PCA chart with coloring/grouping based on different metabolite groups
g1 <- fviz_pca_ind(OC_dat_group_pr, label="none", alpha.ind=1, pointshape=19, habillage=OC_dat_group_cond$Description, pointsize = 3.5, repel = TRUE, col.ind = "black", invisible="quali",legend.title = "Metabolites",
             palette = c("#8b0000", "#d73027", "#f46d43", "#fdae61", "#F0E68C", "#abd9e9", "#74add1", "#4575b4", "#00008b"))

df_edges <- 
  pluck(g1, "data") %>% as_tibble() %>%
  group_by(Groups) %>% 
  summarise(xend = mean(x), yend = mean(y)) %>%
  left_join(y =  pluck(g1, "data"), 
            by = "Groups", 
            multiple = "all")

g1 +
  geom_segment(data = df_edges, aes(xend = xend, yend = yend, x = x, y = y, colour = Groups), alpha = .8, size=1) + 

theme(legend.position = 'right') + coord_equal() + theme_bw() +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()  # Remove minor gridlines
  ) + xlim(-7.5, 7.5) + ylim(-5,5)

# Create PCA chart coloring based on disease association
g2 <- fviz_pca_ind(OC_dat_group_cond_pr, label="none", alpha.ind=1, pointshape=19, habillage=OC_dat_group_cond$Condition, pointsize = 3.5, repel = TRUE, col.ind = "black", invisible="quali",legend.title = "Metabolites",
             palette = c("dodgerblue3", "darkslategray3", "hot pink", "orange"))

df_edges2 <- 
  pluck(g2, "data") %>% as_tibble() %>%
  group_by(Groups) %>% 
  summarise(xend = mean(x), yend = mean(y)) %>%
  left_join(y =  pluck(g2, "data"), 
            by = "Groups", 
            multiple = "all")

g2 +
  geom_segment(data = df_edges2, aes(xend = xend, yend = yend, x = x, y = y, colour = Groups), alpha = .8, size=1) + 

theme(legend.position = 'right') + coord_equal() + theme_bw() +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()  # Remove minor gridlines
  ) + xlim(-7.5, 7.5) + ylim(-5,5)
```
This set of code can be used to save the figures as an EPS file. 

```{r Saving the PCA Charts}
# Saved the PCA charts above into a variable p1, to then export out using the code below:
#cairo_ps(filename='pca_metabolite.eps')
#p1
#dev.off()
```

### 2.3. Box-Whisker Plots
A box-and-whisker plots for each of the individual cytokines are generated here. 

```{r Box-and-Whisker Plots for the Cytokines,fig.height=1.2, fig.width=1.8}
# Remove the "Replicates" number and rename descriptions all into categories based on metabolite name
OC_dat_avg[c(1:3),1] <- "Aminoadipic Acid"
OC_dat_avg[c(4:6),1] <- "Arachidonic Acid"
OC_dat_avg[c(7:9),1] <- "Asparagine"
OC_dat_avg[c(10:12),1] <- "D-Sorbitol"
OC_dat_avg[c(13:15),1] <- "Fructose"
OC_dat_avg[c(16:18),1] <- "L-Tryptophan"
OC_dat_avg[c(19:21),1] <- "Lauric Acid"
OC_dat_avg[c(22:24),1] <- "Niacinamide"
OC_dat_avg[c(25:27),1] <- "Retinol"

# Reorder the groups of metabolites by respective associations with T2D and AD
# AD protective: LT, N;    T2D: LA, Asp;    AD/T2D protective: R;     AD/T2D: AAA, AA, DS, F
OC_boxwhisk_avg <- OC_dat_avg
OC_boxwhisk_avg$Description <- factor(OC_boxwhisk_avg$Description,     # Reorder factor levels
                         levels=c("L-Tryptophan", "Niacinamide", "Lauric Acid", "Asparagine", "Retinol", "Aminoadipic Acid", "Arachidonic Acid", "D-Sorbitol", "Fructose"))

# Box and Whisker Plots - For IL9 color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL9 (33)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + ylim(-2,2.4) +
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL9/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For KC color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `KC (57)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(KC/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For MCP-1 color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `MCP-1 (51)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(MCP-1/Vehicle Ratio)") + ylim(-1.2, 2.0) +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For MIP-1alpha color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `MIP-1alpha (77)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(MIP-1"~alpha~"/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For RANTES color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `RANTES (55)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(RANTES/Vehicle Ratio)") + ylim(-1.5, 1.8) +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For Eotaxin color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `Eotaxin (74)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(Eotaxin/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IFN-gamma color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IFN-gamma (34)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IFN-gamma/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IL-1beta color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL-1Beta (19)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL-1Beta/Vehicle Ratio)") + ylim(-1.1, 3.5) +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IL-1alpha color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL-1alpha (53)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL-1alpha/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IL10 color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL10 (56)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL10/Vehicle Ratio)") + ylim(-1.4,2.3) +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IL12p40 color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL12p40 (76)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL12p40/Vehicle Ratio)") + ylim(-3.3, 2.7) +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IL12p70 color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL12p70 (78)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL12p70/Vehicle Ratio)") + ylim(-2.4, 1.8) + 
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IL13 color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL13 (37)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL13/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For IL6 color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `IL6 (38)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") + ylim(-5.5, 8.1) + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(IL6/Vehicle Ratio)") +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.7, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 

# Box and Whisker Plots - For MIP-1Beta color coded in metabolites and vehicles
ggplot(OC_boxwhisk_avg, aes(x = Description, y = `MIP-1Beta (75)`, fill=Description)) +
    geom_boxplot() + theme_bw() + theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, ), legend.position = "none") + 
    geom_jitter(shape=20, position=position_jitter(0), size=1.5) + xlab("") + ylab(~Log[2]~"(MIP-1Beta/Vehicle Ratio)") + ylim(-4.1, 2.1) +
    scale_fill_manual(breaks = OC_dat$Description, values=c("orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "darkslategray3", "darkslategray3", "darkslategray3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "royalblue3", "royalblue3", "royalblue3", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink", "orange", "orange", "orange", "hot pink", "hot pink", "hot pink")) + geom_hline(yintercept = 0, size = 0.6, linetype="dotted") + geom_hline(yintercept = c(-1,1), size = 0.3, linetype="dashed") 
```

Each box-and-whisker plot can be saved as an EPS file using the code below.

```{r Saving the Box-and-Whisker Plots}
##Saved the charts above into a variable p1, to then export out, used for the box-whisker plots
#cairo_ps(filename='IL-9.eps', width=2.5, height=3.2)
#p1
#dev.off()
```

A Mann-Whitney test is used to determine the p-values across different disease-associations. This is done for each measured cytokine group. The calculated p-values are combined into a single dataframe. 

```{r Mann-Whitney Calculation for p-values}
# Rename a new variable for the Mann-Whitney statistical analysis
OC_MW_dat <- OC_condition_KW

# Renaming the columns to remove the numbers
colnames(OC_MW_dat)[2] <- "Eotaxin"
colnames(OC_MW_dat)[3] <- "IFN-gamma"
colnames(OC_MW_dat)[4] <- "IL-1beta"
colnames(OC_MW_dat)[5] <- "IL-1alpha"
colnames(OC_MW_dat)[6] <- "IL10"
colnames(OC_MW_dat)[7] <- "IL12p40"
colnames(OC_MW_dat)[8] <- "IL12p70"
colnames(OC_MW_dat)[9] <- "IL13"
colnames(OC_MW_dat)[10] <- "IL6"
colnames(OC_MW_dat)[11] <- "IL9"
colnames(OC_MW_dat)[12] <- "KC"
colnames(OC_MW_dat)[13] <- "MCP-1"
colnames(OC_MW_dat)[14] <- "MIP-1beta"
colnames(OC_MW_dat)[15] <- "MIP-1alpha"
colnames(OC_MW_dat)[16] <- "RANTES"

# Establishing four different groups based on disease association for the Mann-Whitney tests
# First establish a group with T2D-associated names only
MW_dat_AD_T2D <- OC_MW_dat[grepl("T2D", OC_MW_dat$Condition),]
# Remove the AD affiliated groups to get T2D-only group
MW_dat_T2D <- MW_dat_AD_T2D[!grepl("AD", MW_dat_AD_T2D$Condition),]
# Get the AD protective group by searching for AD protective
MW_dat_ADprotect <- OC_MW_dat[grepl("AD Protective", OC_MW_dat$Condition),]
# Get the AD-T2D protective by taking the the first group and taking protective
MW_dat_AD_T2Dprotect <- MW_dat_AD_T2D[grepl("Protective", MW_dat_AD_T2D$Condition),]
# Remove the protective association from the AD-T2D harmful group
MW_dat_AD_T2D <- MW_dat_AD_T2D[!grepl("Protective", MW_dat_AD_T2D$Condition),]
# Get the AD association to keep the AD-T2D harmful group
MW_dat_AD_T2D <- MW_dat_AD_T2D[grepl("AD", MW_dat_AD_T2D$Condition),]

# Calculate p-values for each separate cytokine 
# A. Eotaxin Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$Eotaxin, MW_dat_T2D$Eotaxin, alternative = "two.sided")
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$Eotaxin, MW_dat_ADprotect$Eotaxin, alternative = "two.sided")
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$Eotaxin, MW_dat_AD_T2Dprotect$Eotaxin, alternative = "two.sided")
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$Eotaxin, MW_dat_ADprotect$Eotaxin, alternative = "two.sided")
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$Eotaxin, MW_dat_AD_T2Dprotect$Eotaxin, alternative = "two.sided")
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$Eotaxin, MW_dat_AD_T2Dprotect$Eotaxin,alternative = "two.sided")
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
Eotaxin <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for Eotaxin
Eotaxin_MW <- data.frame(Pairwise, Eotaxin)

# B. IFN-gamma Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$'IFN-gamma', MW_dat_T2D$'IFN-gamma', alternative = "two.sided")
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$'IFN-gamma', MW_dat_ADprotect$'IFN-gamma', alternative = "two.sided")
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$'IFN-gamma', MW_dat_AD_T2Dprotect$'IFN-gamma', alternative = "two.sided")
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$'IFN-gamma', MW_dat_ADprotect$'IFN-gamma', alternative = "two.sided")
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$'IFN-gamma', MW_dat_AD_T2Dprotect$'IFN-gamma', alternative = "two.sided")
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$'IFN-gamma', MW_dat_AD_T2Dprotect$'IFN-gamma', alternative = "two.sided")
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IFN_gamma <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for 'IFN-gamma'
IFN_gamma_MW <- data.frame(Pairwise, IFN_gamma)

# C. IL-1Beta Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$'IL-1beta', MW_dat_T2D$'IL-1beta')
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$'IL-1beta', MW_dat_ADprotect$'IL-1beta')
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$'IL-1beta', MW_dat_AD_T2Dprotect$'IL-1beta')
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$'IL-1beta', MW_dat_ADprotect$'IL-1beta')
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$'IL-1beta', MW_dat_AD_T2Dprotect$'IL-1beta')
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$'IL-1beta', MW_dat_AD_T2Dprotect$'IL-1beta')
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL1beta <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for 'IL-1beta'
IL1beta_MW <- data.frame(Pairwise, IL1beta)

# D. IL-1Alpha Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$'IL-1alpha', MW_dat_T2D$'IL-1alpha')
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$'IL-1alpha', MW_dat_ADprotect$'IL-1alpha')
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$'IL-1alpha', MW_dat_AD_T2Dprotect$'IL-1alpha')
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$'IL-1alpha', MW_dat_ADprotect$'IL-1alpha')
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$'IL-1alpha', MW_dat_AD_T2Dprotect$'IL-1alpha')
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$'IL-1alpha', MW_dat_AD_T2Dprotect$'IL-1alpha')
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL1alpha <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for 'IL-1alpha'
IL1alpha_MW <- data.frame(Pairwise, IL1alpha)

# E. IL-10 Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$IL10, MW_dat_T2D$IL10)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$IL10, MW_dat_ADprotect$IL10)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$IL10, MW_dat_AD_T2Dprotect$IL10)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$IL10, MW_dat_ADprotect$IL10)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$IL10, MW_dat_AD_T2Dprotect$IL10)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$IL10, MW_dat_AD_T2Dprotect$IL10)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL10 <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for IL10
IL10_MW <- data.frame(Pairwise, IL10)

# F. IL-12p40 Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$IL12p40, MW_dat_T2D$IL12p40)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$IL12p40, MW_dat_ADprotect$IL12p40)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$IL12p40, MW_dat_AD_T2Dprotect$IL12p40)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$IL12p40, MW_dat_ADprotect$IL12p40)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$IL12p40, MW_dat_AD_T2Dprotect$IL12p40)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$IL12p40, MW_dat_AD_T2Dprotect$IL12p40)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL12p40 <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for IL12p40
IL12p40_MW <- data.frame(Pairwise, IL12p40)

# G. IL-12p70 Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$IL12p70, MW_dat_T2D$IL12p70)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$IL12p70, MW_dat_ADprotect$IL12p70)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$IL12p70, MW_dat_AD_T2Dprotect$IL12p70)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$IL12p70, MW_dat_ADprotect$IL12p70)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$IL12p70, MW_dat_AD_T2Dprotect$IL12p70)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$IL12p70, MW_dat_AD_T2Dprotect$IL12p70)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL12p70 <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for IL12p70
IL12p70_MW <- data.frame(Pairwise, IL12p70)

# H. IL-13 Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$IL13, MW_dat_T2D$IL13)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$IL13, MW_dat_ADprotect$IL13)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$IL13, MW_dat_AD_T2Dprotect$IL13)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$IL13, MW_dat_ADprotect$IL13)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$IL13, MW_dat_AD_T2Dprotect$IL13)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$IL13, MW_dat_AD_T2Dprotect$IL13)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL13 <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for IL13
IL13_MW <- data.frame(Pairwise, IL13)

# I. IL-6 Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$IL6, MW_dat_T2D$IL6)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$IL6, MW_dat_ADprotect$IL6)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$IL6, MW_dat_AD_T2Dprotect$IL6)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$IL6, MW_dat_ADprotect$IL6)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$IL6, MW_dat_AD_T2Dprotect$IL6)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$IL6, MW_dat_AD_T2Dprotect$IL6)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL6 <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for IL6
IL6_MW <- data.frame(Pairwise, IL6)

# J. IL-9 Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$IL9, MW_dat_T2D$IL9)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$IL9, MW_dat_ADprotect$IL9)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$IL9, MW_dat_AD_T2Dprotect$IL9)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$IL9, MW_dat_ADprotect$IL9)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$IL9, MW_dat_AD_T2Dprotect$IL9)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$IL9, MW_dat_AD_T2Dprotect$IL9)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
IL9 <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for IL9
IL9_MW <- data.frame(Pairwise, IL9)

# K. KC Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$KC, MW_dat_T2D$KC)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$KC, MW_dat_ADprotect$KC)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$KC, MW_dat_AD_T2Dprotect$KC)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$KC, MW_dat_ADprotect$KC)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$KC, MW_dat_AD_T2Dprotect$KC)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$KC, MW_dat_AD_T2Dprotect$KC)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
KC <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for KC
KC_MW <- data.frame(Pairwise, KC)

# L. MCP-1 Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$'MCP-1', MW_dat_T2D$'MCP-1')
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$'MCP-1', MW_dat_ADprotect$'MCP-1')
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$'MCP-1', MW_dat_AD_T2Dprotect$'MCP-1')
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$'MCP-1', MW_dat_ADprotect$'MCP-1')
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$'MCP-1', MW_dat_AD_T2Dprotect$'MCP-1')
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$'MCP-1', MW_dat_AD_T2Dprotect$'MCP-1')
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
MCP1 <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for 'MCP-1'
MCP1_MW <- data.frame(Pairwise, MCP1)

# M. MIP-1beta Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$'MIP-1beta', MW_dat_T2D$'MIP-1beta')
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$'MIP-1beta', MW_dat_ADprotect$'MIP-1beta')
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$'MIP-1beta', MW_dat_AD_T2Dprotect$'MIP-1beta')
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$'MIP-1beta', MW_dat_ADprotect$'MIP-1beta')
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$'MIP-1beta', MW_dat_AD_T2Dprotect$'MIP-1beta')
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$'MIP-1beta', MW_dat_AD_T2Dprotect$'MIP-1beta')
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
MIP1beta <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for 'MIP-1beta'
MIP1beta_MW <- data.frame(Pairwise, MIP1beta)

# O. MIP-1alpha Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$'MIP-1alpha', MW_dat_T2D$'MIP-1alpha')
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$'MIP-1alpha', MW_dat_ADprotect$'MIP-1alpha')
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$'MIP-1alpha', MW_dat_AD_T2Dprotect$'MIP-1alpha')
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$'MIP-1alpha', MW_dat_ADprotect$'MIP-1alpha')
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$'MIP-1alpha', MW_dat_AD_T2Dprotect$'MIP-1alpha')
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$'MIP-1alpha', MW_dat_AD_T2Dprotect$'MIP-1alpha')
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
MIP1alpha <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for 'MIP-1alpha'
MIP1alpha_MW <- data.frame(Pairwise, MIP1alpha)

# P. RANTES Cytokine Mann Whitney Comparison
# AD-T2D harmful vs T2D harmful
AD_T2DvsT2D <- wilcox.test(MW_dat_AD_T2D$RANTES, MW_dat_T2D$RANTES)
# AD-T2D harmful vs AD protective
AD_T2DvsADprotect <- wilcox.test(MW_dat_AD_T2D$RANTES, MW_dat_ADprotect$RANTES)
# AD-T2D harmful vs AD-T2D protective
AD_T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_AD_T2D$RANTES, MW_dat_AD_T2Dprotect$RANTES)
# T2D harmful vs AD protective
T2DvsADprotect <- wilcox.test(MW_dat_T2D$RANTES, MW_dat_ADprotect$RANTES)
# T2D harmful vs AD-T2D protective
T2DvsAD_T2Dprotect <- wilcox.test(MW_dat_T2D$RANTES, MW_dat_AD_T2Dprotect$RANTES)
# AD Protective vs AD-T2D protective
ADprotectvsAD_T2Dprotect <- wilcox.test(MW_dat_ADprotect$RANTES, MW_dat_AD_T2Dprotect$RANTES)
# Create a list for the disease-associated pair relationships 
Pairwise <- c("AD-T2D vs T2D", "AD-T2D vs AD protective", "AD-T2D vs AD-T2D protective", "T2D vs AD protective", "T2D vs AD-T2D protective", "AD protective vs AD-T2D protective")
# Prepare the pvalues in a list
RANTES <- c(AD_T2DvsT2D$p.value, AD_T2DvsADprotect$p.value, AD_T2DvsAD_T2Dprotect$p.value, T2DvsADprotect$p.value, T2DvsAD_T2Dprotect$p.value, ADprotectvsAD_T2Dprotect$p.value)
# Generate Data frame for the MW test for RANTES
RANTES_MW <- data.frame(Pairwise, RANTES)

# Combine the p-values from Mann-Whitney into one data frame
Cytokines_MW <- data.frame(Pairwise, Eotaxin_MW$Eotaxin, IFN_gamma_MW$IFN_gamma, IL1beta_MW$IL1beta, IL1alpha_MW$IL1alpha, IL10_MW$IL10, IL12p40_MW$IL12p40, IL12p70_MW$IL12p70, IL13_MW$IL13, IL6_MW$IL6, IL9_MW$IL9, KC_MW$KC, MCP1_MW$MCP1, MIP1beta_MW$MIP1beta, MIP1alpha_MW$MIP1alpha, RANTES_MW$RANTES)
```

### 2.4. Partial Least Squares Discriminant Analysis (PLS-DA)
PLS-DA model using a four-way analysis among the different disease-associated metabolites.

#### 2.4.1. PLS-DA for 4-Way Comparison Based on Disease-Association
```{r Prepare the Data for a 4-Way PLS-DA Comparison, fig.height=4.2, fig.width=5}
# Rename a new variable for the plsda data set
OC_plsda_dat <- OC_condition_KW

# Renaming the columns to remove the numbers
colnames(OC_plsda_dat)[2] <- "Eotaxin"
colnames(OC_plsda_dat)[3] <- "IFN-gamma"
colnames(OC_plsda_dat)[4] <- "IL-1beta"
colnames(OC_plsda_dat)[5] <- "IL-1alpha"
colnames(OC_plsda_dat)[6] <- "IL10"
colnames(OC_plsda_dat)[7] <- "IL12p40"
colnames(OC_plsda_dat)[8] <- "IL12p70"
colnames(OC_plsda_dat)[9] <- "IL13"
colnames(OC_plsda_dat)[10] <- "IL6"
colnames(OC_plsda_dat)[11] <- "IL9"
colnames(OC_plsda_dat)[12] <- "KC"
colnames(OC_plsda_dat)[13] <- "MCP-1"
colnames(OC_plsda_dat)[14] <- "MIP-1beta"
colnames(OC_plsda_dat)[15] <- "MIP-1alpha"
colnames(OC_plsda_dat)[16] <- "RANTES"

# Factorize the conditions based on the four different disease-association groups
OC_plsda_dat$Condition <- factor(OC_plsda_dat$Condition)

## Cross Validation to Determine the Number of Components
# Separate data into the specific categories (Y, disease association) and (X, dataset from cytokines)
Y <- as.character(OC_plsda_dat$Condition)
X <- OC_plsda_dat[2:16]

# Compute PLSDA for 10 components (noted as plsda_draft for multiple components)
plsda_draft <- plsda(X, Y, ncomp = 10)

# Perform a 3-fold cross-validation repeated 10 times to evaluate classification performance
# Noted as plsda_cv for the cross-validation analysis
plsda_cv <- perf(plsda_draft, validation = 'Mfold', folds = 3, 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 100)
# Plot the balanced error rate (BER) and the overall error
plot(plsda_cv, sd = TRUE, legend.position = 'horizontal')

## Plotting the PLS-DA Graph
# Code below is used to save the PLS-DA file
#cairo_ps(filename='plsda_4way.eps', height = 8.2, width = 10)

# A four-way PLS-DA was performed with 3 components
plsda_4way <- plsda(X, Y, ncomp = 3)

# PLS-DA with LV1 and LV2
plotIndiv(plsda_4way, ind.names = FALSE, legend=TRUE,
          comp=c(1,2), ellipse = FALSE, pch=16,
          title = "",
          star = TRUE, 
          size.title = 0.1,
          size.legend = 0.5,
          style="graphics",
          cex = 3.5,
          point.lwd = 4.5,
          X.label = 'Scores on LV1', Y.label = 'Scores on LV2',
          col = c("hot pink", "darkslategray3", "royalblue3", "orange"))

          abline(h=0, lty="dotted", col="black", lwd=2)
          abline(v=0, lty="dotted", col="black", lwd=2)

#dev.off()

# Code is used to identify which metabolite is contributing the the data point in the PLS-DA Model
# Identify position in the row to identify metabolite association
plotIndiv(plsda_4way, ind.names = TRUE, legend=TRUE,
          comp=c(1,2), ellipse = FALSE,
          title = "",
          star = TRUE, 
          size.title = 0.1,
          size.legend = 0.5,
          style="graphics",
          cex = 3.5,
          point.lwd = 4.5,
          X.label = 'Scores on LV1', Y.label = 'Scores on LV2',
          col = c("hot pink", "darkslategray3", "royalblue3", "orange"))

          abline(h=0, lty="dotted", col="black", lwd=2)
          abline(v=0, lty="dotted", col="black", lwd=2)
```

Identify the loadings for latent variable 1 and 2. Determine the variable importance in projection for the latent variables 1 and 2.

```{r Loading Variables and VIP Determination,  fig.height=5.2, fig.width=10}
## Determine the Variable Importance in Projections for the Loadings
# Determine the VIP values for PLSDA-4way
plsda_4way.vip <- vip(plsda_4way)

barplot(plsda_4way.vip,
        beside = TRUE, col = c("red", "orange", "yellow", "dark green", "light blue", "purple", "violet", "brown", "pink", "black", "white", "gray", "blue", "light green", "lightcyan"),
        ylim = c(0, 2.0), legend = rownames(plsda_4way.vip),
        main = "Variable Importance in the Projection", font.main = 4)

abline(h=1, lty="dotted", col="black", lwd=5)

## Latent Variable 1 Loadings
# Pull the loadings information
loadings4way <- as.data.frame(plsda_4way[["loadings"]][["X"]])
# Capture the rowname list
rowname4way <- rownames(loadings4way)
# Remove the old dataframe's row name
rownames(loadings4way) <- NULL
# Merge the row name (as a first column) with the other loadings
loadings4way <- cbind(rowname4way, loadings4way)
# Rename the first column to Cytokine
colnames(loadings4way)[1] <- "Cytokine"

# Specify the desired words in the desired order
selected_words <- c("IL-1alpha", "MIP-1alpha","IL12p70", "IL6", "IL13", "IL-1beta", "KC", "Eotaxin",  "IL10", "MCP-1", "IFN-gamma", "IL12p40", "RANTES", "MIP-1beta", "IL9")

# Rearrange the rows based on selected words
loadings4way <- loadings4way[match(selected_words, loadings4way$Cytokine, nomatch = 0), ]
rownames(loadings4way) <- NULL

# Preparing factor to establish the barplot positioning in the figure across all groups
loadings4way$Cytokine <- factor(loadings4way$Cytokine,
                  levels = c("IL9", "MIP-1beta", "RANTES", "IL12p40", "IFN-gamma", "MCP-1", "IL10", "Eotaxin", "KC", "IL-1beta", "IL13", "IL6", "IL12p70", "MIP-1alpha", "IL-1alpha"))

# Plot loadings with each variable identified with highest contribution to a specific class
plotLoadings(plsda_4way, comp = 1, method = 'mean', contrib = 'max', border = "black", xlim = c(-0.3,0.55))
# Plot of loadings color adjusted with fixed cytokine list
loadings4way %>%
        ggplot(aes(x = Cytokine, y = comp1, fill = Cytokine)) +
            geom_bar(stat = "identity", color="black", show.legend = FALSE, width = 1) +
            scale_fill_manual(values = c("orange", "royalblue3", "royalblue3", "orange", "orange", "darkslategray3", "orange", "orange", "darkslategray3", "orange", "orange", "darkslategray3", "royalblue3", "royalblue3", "hot pink")) + 
            coord_flip() + theme_classic() + labs(x=NULL, y="Loadings on LV1") +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  )

## Latent Variable 2 Loadings
# Pull the loadings information
loadings4way <- as.data.frame(plsda_4way[["loadings"]][["X"]])
# Capture the rowname list
rowname4way <- rownames(loadings4way)
# Remove the old dataframe's row name
rownames(loadings4way) <- NULL
# Merge the row name (as a first column) with the other loadings
loadings4way <- cbind(rowname4way, loadings4way)
# Rename the first column to Cytokine
colnames(loadings4way)[1] <- "Cytokine"

# Specify the desired words in the desired order
selected_words <- c("RANTES", "IFN-gamma","MIP-1alpha", "KC", "IL9", "IL13", "Eotaxin", "MIP-1beta",  "IL-1alpha", "IL6", "IL-1beta", "IL10", "IL12p40", "IL12p70", "MCP-1")

# Rearrange the rows based on selected words
loadings4way <- loadings4way[match(selected_words, loadings4way$Cytokine, nomatch = 0), ]
rownames(loadings4way) <- NULL

# Preparing factor to establish the barplot positioning in the figure across all groups
loadings4way$Cytokine <- factor(loadings4way$Cytokine,
                  levels = c("MCP-1", "IL12p70", "IL12p40", "IL10", "IL-1beta", "IL6", "IL-1alpha", "MIP-1beta", "Eotaxin", "IL13", "IL9", "KC", "MIP-1alpha", "IFN-gamma", "RANTES"))

# Plot loadings with each variable identified with highest contribution to a specific class
plotLoadings(plsda_4way, comp = 2, method = 'mean', contrib = 'max')
# Plot of loadings color adjusted with fixed cytokine list
loadings4way %>%
        ggplot(aes(x = Cytokine, y = comp2, fill = Cytokine)) +
            geom_bar(stat = "identity", color="black", show.legend = FALSE, width = 1) +
            scale_fill_manual(values = c("darkslategray3", "royalblue3", "orange", "orange", "orange", "darkslategray3", "hot pink", "royalblue3", "orange", "orange", "orange", "darkslategray3", "royalblue3", "orange", "royalblue3")) + 
            coord_flip() + theme_classic() + labs(x=NULL, y="Loadings on LV2") +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()  ,# Remove minor gridlines
  )
```

#### 2.4.2. PLS-DA for 2-Way Comparison Based on AD Harmful vs. AD Protective
Prepare a similar PLS-DA model, including only the metabolites and data associated with AD.

```{r Prepare the Data for a 2-Way AD PLS-DA Comparison, fig.height=4.2, fig.width=5}
# Rename a new variable for the plsda data set
OC_plsda_dat_AD <- OC_condition_KW

# Renaming the columns to remove the numbers
colnames(OC_plsda_dat_AD)[2] <- "Eotaxin"
colnames(OC_plsda_dat_AD)[3] <- "IFN-gamma"
colnames(OC_plsda_dat_AD)[4] <- "IL-1beta"
colnames(OC_plsda_dat_AD)[5] <- "IL-1alpha"
colnames(OC_plsda_dat_AD)[6] <- "IL10"
colnames(OC_plsda_dat_AD)[7] <- "IL12p40"
colnames(OC_plsda_dat_AD)[8] <- "IL12p70"
colnames(OC_plsda_dat_AD)[9] <- "IL13"
colnames(OC_plsda_dat_AD)[10] <- "IL6"
colnames(OC_plsda_dat_AD)[11] <- "IL9"
colnames(OC_plsda_dat_AD)[12] <- "KC"
colnames(OC_plsda_dat_AD)[13] <- "MCP-1"
colnames(OC_plsda_dat_AD)[14] <- "MIP-1beta"
colnames(OC_plsda_dat_AD)[15] <- "MIP-1alpha"
colnames(OC_plsda_dat_AD)[16] <- "RANTES"

# Adjust disease-association on AD protective or harmful, regardless of T2D status
OC_plsda_dat_AD[c(1:3),1] <- "AD"
OC_plsda_dat_AD[c(4:6),1] <- "AD"
OC_plsda_dat_AD[c(7:9),1] <- "T2D"
OC_plsda_dat_AD[c(10:12),1] <- "AD"
OC_plsda_dat_AD[c(13:15),1] <- "AD"
OC_plsda_dat_AD[c(16:18),1] <- "AD Protective"
OC_plsda_dat_AD[c(19:21),1] <- "T2D"
OC_plsda_dat_AD[c(22:24),1] <- "AD Protective"
OC_plsda_dat_AD[c(25:27),1] <- "AD Protective"

# Select only the categories associated with AD, and NOT T2D
OC_plsda_dat_AD <- OC_plsda_dat_AD[grepl("AD", OC_plsda_dat_AD$Condition),]

# Factorize the conditions based on association to AD protective or harmful, regardless of T2D affiliation
OC_plsda_dat$Condition <- factor(OC_plsda_dat$Condition)

## Cross Validation to determine the number of components for the model
# Separate data into the specific categories (Y, disease association) and (X, dataset from cytokines)
Y_AD <- as.character(OC_plsda_dat_AD$Condition)
X_AD <- OC_plsda_dat_AD[2:16]

# Compute PLSDA for 10 components (noted as plsda_draft for multiple components)
plsda_draft_AD <- plsda(X_AD, Y_AD, ncomp = 10)

# Perform a 3-fold cross-validation repeated 10 times to evaluate classification performance
# Noted as plsda_cv for the cross-validation analysis
plsda_cv_AD <- perf(plsda_draft_AD, validation = 'Mfold', folds = 3, 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 100)
# Plot the balanced error rate (BER) and the overall error
plot(plsda_cv_AD, sd = TRUE, legend.position = 'horizontal')

## Plot the PLS-DA Model
# Code below is to save the figure as an EPS file
#cairo_ps(filename='plsda_AD_2way.eps', height = 8.2, width = 10)

# A two-way PLS-DA was performed with 2 components
plsda_AD <- plsda(X_AD, Y_AD, ncomp = 2)

# PLS-DA with LV1 and LV2
plotIndiv(plsda_AD, ind.names = FALSE, legend=TRUE,
          comp=c(1,2), ellipse = FALSE, pch=16,
          title = "",
          star = TRUE, 
          size.title = 0.1,
          size.legend = 0.5,
          style="graphics",
          cex = 3.5,
          point.lwd = 4.5,
          X.label = 'Scores on LV1', Y.label = 'Scores on LV2',
          col = c("red3", "royalblue3"))
          
          abline(h=0, lty="dotted", col="black", lwd=2)
          abline(v=0, lty="dotted", col="black", lwd=2)
          
#dev.off()

## PLS-DA Model with the associated number to identify the metabolite on the figure
# Identify position in the row to identify metabolite association
plotIndiv(plsda_AD, ind.names = TRUE, legend=TRUE,
          comp=c(1,2), ellipse = FALSE,
          title = "",
          star = TRUE, 
          size.title = 0.1,
          size.legend = 0.5,
          style="graphics",
          cex = 3.5,
          point.lwd = 4.5,
          X.label = 'Scores on LV1', Y.label = 'Scores on LV2',
          col = c("red3", "royalblue3"))

          abline(h=0, lty="dotted", col="black", lwd=2)
          abline(v=0, lty="dotted", col="black", lwd=2)
```

Identify the loadings for latent variable 1 and 2. Determine the variable importance in projection for the latent variables 1 and 2.

```{r Loading Variables and VIP Determination for AD,  fig.height=5.2, fig.width=10}
# Determine the VIP values for AD-2way 
plsda_AD.vip <- vip(plsda_AD)

# Plot the graph for the VIP>1
barplot(plsda_AD.vip,
        beside = TRUE, col = c("red", "orange", "yellow", "dark green", "light blue", "purple", "violet", "brown", "pink", "black", "white", "gray", "blue", "light green", "lightcyan"),
        ylim = c(0, 2.0), legend = rownames(plsda_AD.vip),
        main = "Variable Importance in the Projection", font.main = 4)

abline(h=1, lty="dotted", col="black", lwd=5)

## Loadings for Latent Variable 1
# Pull the loadings information
loadingsAD <- as.data.frame(plsda_AD[["loadings"]][["X"]])
# Capture the rowname list
rownameAD <- rownames(loadingsAD)
# Remove the old dataframe's row name
rownames(loadingsAD) <- NULL
# Merge the row name (as a first column) with the other loadings
loadingsAD <- cbind(rownameAD, loadingsAD)
# Rename the first column to Cytokine
colnames(loadingsAD)[1] <- "Cytokine"

# Specify the desired words in the desired order
selected_words <- c("IFN-gamma", "Eotaxin","MIP-1alpha", "MIP-1beta", "RANTES", "IL13", "IL9", "IL6", "IL12p70", "KC", "IL-1beta", "IL-1alpha", "IL12p40", "IL10", "MCP-1")

# Rearrange the rows based on selected words
loadingsAD <- loadingsAD[match(selected_words, loadingsAD$Cytokine, nomatch = 0), ]
rownames(loadingsAD) <- NULL

# Preparing factor to establish the barplot positioning in the figure across all groups
loadingsAD$Cytokine <- factor(loadingsAD$Cytokine,
                  levels = c("MCP-1", "IL10", "IL12p40", "IL-1alpha", "IL-1beta", "KC", "IL12p70", "IL6", "IL9", "IL13", "RANTES", "MIP-1beta", "MIP-1alpha", "Eotaxin", "IFN-gamma"))

# Plot loadings with each variable identified with highest contribution to a specific class
plotLoadings(plsda_AD, comp = 1, method = 'mean', contrib = 'max')
# Plot of loadings color adjusted with fixed cytokine list
loadingsAD %>%
        ggplot(aes(x = Cytokine, y = comp1, fill = Cytokine)) +
            geom_bar(stat = "identity", color="black", show.legend = FALSE, width = 1) +
            scale_fill_manual(values = c("royalblue3", "royalblue3", "royalblue3", "red3", "royalblue3", "royalblue3", "red3", "royalblue3", "red3", "royalblue3", "royalblue3", "red3", "royalblue3", "royalblue3", "red3")) + 
            coord_flip() + theme_classic() + labs(x=NULL, y="Loadings on LV1") +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  )

## Loadings for Latent Variable 2
# Pull the loadings information
loadingsAD <- as.data.frame(plsda_AD[["loadings"]][["X"]])
# Capture the rowname list
rownameAD <- rownames(loadingsAD)
# Remove the old dataframe's row name
rownames(loadingsAD) <- NULL
# Merge the row name (as a first column) with the other loadings
loadingsAD <- cbind(rownameAD, loadingsAD)
# Rename the first column to Cytokine
colnames(loadingsAD)[1] <- "Cytokine"

# Specify the desired words in the desired order
selected_words <- c("IL9", "IL13", "IL-1beta", "MIP-1beta", "IL12p40", "RANTES", "MCP-1", "IL10", "IFN-gamma", "KC", "IL6", "Eotaxin", "IL12p70", "IL-1alpha", "MIP-1alpha")

# Rearrange the rows based on selected words
loadingsAD <- loadingsAD[match(selected_words, loadingsAD$Cytokine, nomatch = 0), ]
rownames(loadingsAD) <- NULL

# Preparing factor to establish the barplot positioning in the figure across all groups
loadingsAD$Cytokine <- factor(loadingsAD$Cytokine,
                  levels = c("MIP-1alpha", "IL-1alpha", "IL12p70", "Eotaxin", "IL6", "KC", "IFN-gamma", "IL10", "MCP-1", "RANTES", "IL12p40", "MIP-1beta", "IL-1beta", "IL13", "IL9"))

# Plot loadings with each variable identified with highest contribution to a specific class
plotLoadings(plsda_AD, comp = 2, method = 'mean', contrib = 'max')

# Plot of loadings color adjusted with fixed cytokine list
loadingsAD %>%
        ggplot(aes(x = Cytokine, y = comp2, fill = Cytokine)) +
            geom_bar(stat = "identity", color="black", show.legend = FALSE, width = 1) +
            scale_fill_manual(values = c("royalblue3", "red3", "red3", "royalblue3", "royalblue3", "royalblue3", "red3", "royalblue3", "royalblue3", "royalblue3", "royalblue3", "red3", "royalblue3", "royalblue3", "red3")) + 
            coord_flip() + theme_classic() + labs(x=NULL, y="Loadings on LV2") +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  )
```

#### 2.4.3. PLS-DA for 2-Way Comparison Based on T2D Harmful vs. T2D Protective
Prepare a similar PLS-DA model, including only the metabolites and data associated with T2D.

```{r Prepare the Data for a 2-Way T2D PLS-DA Comparison, fig.height=4.2, fig.width=5}
# Rename a new variable for the plsda data set
OC_plsda_dat_T2D <- OC_condition_KW

# Renaming the columns to remove the numbers
colnames(OC_plsda_dat_T2D)[2] <- "Eotaxin"
colnames(OC_plsda_dat_T2D)[3] <- "IFN-gamma"
colnames(OC_plsda_dat_T2D)[4] <- "IL-1beta"
colnames(OC_plsda_dat_T2D)[5] <- "IL-1alpha"
colnames(OC_plsda_dat_T2D)[6] <- "IL10"
colnames(OC_plsda_dat_T2D)[7] <- "IL12p40"
colnames(OC_plsda_dat_T2D)[8] <- "IL12p70"
colnames(OC_plsda_dat_T2D)[9] <- "IL13"
colnames(OC_plsda_dat_T2D)[10] <- "IL6"
colnames(OC_plsda_dat_T2D)[11] <- "IL9"
colnames(OC_plsda_dat_T2D)[12] <- "KC"
colnames(OC_plsda_dat_T2D)[13] <- "MCP-1"
colnames(OC_plsda_dat_T2D)[14] <- "MIP-1beta"
colnames(OC_plsda_dat_T2D)[15] <- "MIP-1alpha"
colnames(OC_plsda_dat_T2D)[16] <- "RANTES"

# Adjust disease-association on T2D protective or harmful, regardless of T2D status
OC_plsda_dat_T2D[c(1:3),1] <- "T2D"
OC_plsda_dat_T2D[c(4:6),1] <- "T2D"
OC_plsda_dat_T2D[c(7:9),1] <- "T2D"
OC_plsda_dat_T2D[c(10:12),1] <- "T2D"
OC_plsda_dat_T2D[c(13:15),1] <- "T2D"
OC_plsda_dat_T2D[c(16:18),1] <- "AD Protective"
OC_plsda_dat_T2D[c(19:21),1] <- "T2D"
OC_plsda_dat_T2D[c(22:24),1] <- "AD Protective"
OC_plsda_dat_T2D[c(25:27),1] <- "T2D Protective"

# Select only the categories associated with T2D, and NOT T2D
OC_plsda_dat_T2D <- OC_plsda_dat_T2D[grepl("T2D", OC_plsda_dat_T2D$Condition),]

# Factorize the conditions based on association to T2D protective or harmful, regardless of T2D affiliation
OC_plsda_dat$Condition <- factor(OC_plsda_dat$Condition)

## Cross Validation to determine the number of components for the model
# Separate data into the specific categories (Y, disease association) and (X, dataset from cytokines)
Y_T2D <- as.character(OC_plsda_dat_T2D$Condition)
X_T2D <- OC_plsda_dat_T2D[2:16]

# Compute PLSDA for 10 components (noted as plsda_draft for multiple components)
plsda_draft_T2D <- plsda(X_T2D, Y_T2D, ncomp = 10)

# Perform a 3-fold cross-validation repeated 10 times to evaluate classification performance
# Noted as plsda_cv for the cross-validation analysis
plsda_cv_T2D <- perf(plsda_draft_T2D, validation = 'Mfold', folds = 3, 
                  progressBar = FALSE,  # Set to TRUE to track progress
                  nrepeat = 100)
# Plot the balanced error rate (BER) and the overall error
plot(plsda_cv_T2D, sd = TRUE, legend.position = 'horizontal')

## PLS-DA Model Being Plotted
# Code below to save the Figure as an EPS
#cairo_ps(filename='plsda_T2D_2way.eps', height = 8.2, width = 10)

# A two-way T2D PLS-DA was performed with 3 components
plsda_T2D <- plsda(X_T2D, Y_T2D, ncomp = 3)

# PLS-DA with LV1 and LV2
plotIndiv(plsda_T2D, ind.names = FALSE, legend=TRUE,
          comp=c(1,2), ellipse = FALSE, pch=16,
          title = "",
          star = TRUE, 
          size.title = 0.1,
          size.legend = 0.5,
          style="graphics",
          cex = 3.5,
          point.lwd = 4.5,
          X.label = 'Scores on LV1', Y.label = 'Scores on LV2',
          col = c("plum", "gray60"))

          abline(h=0, lty="dotted", col="black", lwd=2)
          abline(v=0, lty="dotted", col="black", lwd=2)
          
#dev.off()

## PLS-DA Model to identify the individual metabolites on the figure
# Identify position in the row to identify metabolite association
plotIndiv(plsda_T2D, ind.names = TRUE, legend=TRUE,
          comp=c(1,2), ellipse = FALSE,
          title = "",
          star = TRUE, 
          size.title = 0.1,
          size.legend = 0.5,
          style="graphics",
          cex = 3.5,
          point.lwd = 4.5,
          X.label = 'Scores on LV1', Y.label = 'Scores on LV2',
          col = c("plum", "gray60"))

          abline(h=0, lty="dotted", col="black", lwd=2)
          abline(v=0, lty="dotted", col="black", lwd=2)
```

Identify the loadings for latent variable 1 and 2. Determine the variable importance in projection for the latent variables 1 and 2.

```{r Loading Variables and VIP Determination for T2D,  fig.height=5.2, fig.width=10}
# Determine the VIP values for PLSDA-4way
plsda_T2D.vip <- vip(plsda_T2D)

barplot(plsda_T2D.vip,
        beside = TRUE, col = c("red", "orange", "yellow", "dark green", "light blue", "purple", "violet", "brown", "pink", "black", "white", "gray", "blue", "light green", "lightcyan"),
        ylim = c(0, 2.0), legend = rownames(plsda_T2D.vip),
        main = "Variable Importance in the Projection", font.main = 4)

abline(h=1, lty="dotted", col="black", lwd=5)

## Loadings for Latent Variable 1
# Pull the loadings information
loadingsT2D <- as.data.frame(plsda_T2D[["loadings"]][["X"]])
# Capture the rowname list
rownameT2D <- rownames(loadingsT2D)
# Remove the old dataframe's row name
rownames(loadingsT2D) <- NULL
# Merge the row name (as a first column) with the other loadings
loadingsT2D <- cbind(rownameT2D, loadingsT2D)
# Rename the first column to Cytokine
colnames(loadingsT2D)[1] <- "Cytokine"

# Specify the desired words in the desired order
selected_words <- c("IL13", "IL-1beta", "IL12p70", "RANTES", "Eotaxin", "MIP-1alpha", "IL6", "IFN-gamma", "IL10", "IL12p40", "MIP-1beta", "IL9", "IL-1alpha", "KC", "MCP-1")

# Rearrange the rows based on selected words
loadingsT2D <- loadingsT2D[match(selected_words, loadingsT2D$Cytokine, nomatch = 0), ]
rownames(loadingsT2D) <- NULL

# Preparing factor to establish the barplot positioning in the figure across all groups
loadingsT2D$Cytokine <- factor(loadingsT2D$Cytokine,
                  levels = c("MCP-1", "MIP-1beta", "IL9", "RANTES", "IL6", "IL12p70", "KC", "IFN-gamma", "IL-1alpha", "Eotaxin", "IL13", "IL12p40", "MIP-1alpha", "IL-1beta", "IL10"))

# Plot loadings with each variable identified with highest contribution to a specific class
plotLoadings(plsda_T2D, comp = 1, method = 'mean', contrib = 'max')
# Plot of loadings color adjusted with fixed cytokine list
loadingsT2D %>%
        ggplot(aes(x = Cytokine, y = comp1, fill = Cytokine)) +
            geom_bar(stat = "identity", color="black", show.legend = FALSE, width = 1) +
            scale_fill_manual(values = c("gray60", "plum", "plum", "plum", "gray60", "plum", "gray60", "plum", "plum", "plum", "plum", "plum", "plum", "gray60", "gray60")) + 
            coord_flip() + theme_classic() + labs(x=NULL, y="Loadings on LV1") +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  )

## Loadings for Latent Variable 2
# Pull the loadings information
loadingsT2D <- as.data.frame(plsda_T2D[["loadings"]][["X"]])
# Capture the rowname list
rownameT2D <- rownames(loadingsT2D)
# Remove the old dataframe's row name
rownames(loadingsT2D) <- NULL
# Merge the row name (as a first column) with the other loadings
loadingsT2D <- cbind(rownameT2D, loadingsT2D)
# Rename the first column to Cytokine
colnames(loadingsT2D)[1] <- "Cytokine"

# Specify the desired words in the desired order
selected_words <- c("IL10", "IL-1beta", "IL12p40", "Eotaxin", "IL13", "IL9", "IFN-gamma", "MCP-1", "IL6", "MIP-1alpha", "KC", "IL-1alpha", "MIP-1beta", "IL12p70", "RANTES")

# Rearrange the rows based on selected words
loadingsT2D <- loadingsT2D[match(selected_words, loadingsT2D$Cytokine, nomatch = 0), ]
rownames(loadingsT2D) <- NULL

# Preparing factor to establish the barplot positioning in the figure across all groups
loadingsT2D$Cytokine <- factor(loadingsT2D$Cytokine,
                  levels = c("MIP-1alpha", "MCP-1", "IL10", "RANTES", "IFN-gamma", "KC", "IL12p70", "IL-1beta", "IL-1alpha", "IL12p40", "IL13", "Eotaxin", "IL9", "MIP-1beta", "IL6"))

# Plot loadings with each variable identified with highest contribution to a specific class
plotLoadings(plsda_T2D, comp = 2, method = 'mean', contrib = 'max')
# Plot of loadings color adjusted with fixed cytokine list
loadingsT2D %>%
        ggplot(aes(x = Cytokine, y = comp2, fill = Cytokine)) +
            geom_bar(stat = "identity", color="black", show.legend = FALSE, width = 1) +
            scale_fill_manual(values = c("plum","gray60", "gray60", "plum", "plum", "gray60", "plum", "gray60", "plum", "plum", "plum", "plum", "plum", "plum", "gray60")) + 
            coord_flip() + theme_classic() + labs(x=NULL, y="Loadings on LV2") +
  theme(
    panel.grid.major = element_blank(),  # Remove major gridlines
    panel.grid.minor = element_blank()   # Remove minor gridlines
  )
```

Saving the PLS-DA Model and Figures as an EPS file

```{r Saving the PLS-DA Figures}
#Saved the charts above into a variable p1, to then export out
#cairo_ps(filename='PLSDA-T2D-LV1.eps')
#p1
#dev.off()
```


*** END OF SCRIPT ***
